# Reproducible Research - Project 1

## Xiao Yu

### Examining data from a personal activity monitoring device 

==============================================================

## Loading and Preprocessing the Data

```{r chunk for loading and reading data, results="hide"}
## load packages and set work directory
setwd("/Users/yxshelly/Desktop/data_science/c5_reproducible research")
library(dplyr); library(plyr); library(ggplot2)
setwd("/Users/yxshelly/Desktop/data_science/c5_reproducible research")
if(!file.exists("data")) {
        dir.create("data")
}
## download data from the website to "data"
path <- getwd()
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
              destfile = paste(path, "data/files.zip", sep = "/"))
list.files("./data")
unzip(zipfile = "data/files.zip")
# document the date downloaded
dataDownloaded <- date()
```

### load the data and briefly examine
```{r loading data} 
activity <-read.csv("activity.csv")
str(activity)
head(activity)
```

===========================================================================

## What is mean total number of steps taken per day?

### 1. Calculate the total number of steps taken per day

```{r}
totalSteps <- ddply(activity, "date", summarise, tot=sum(steps), na.rm=T)
head(totalSteps)
```

### 2. Make a histogram of the total number of steps taken each day

```{r plotting saving to figure folder, results= "hide", echo=FALSE}
jpeg("figure/figure1.jpeg", height=600, width=800, quality=8000)
ggplot(totalSteps, aes(x=tot)) +
        geom_histogram(fill= "red", binwidth=1000) +
        labs(title = "Histogram of Total Number of Daily Steps", x = "Steps", y = "Frequency")
dev.off()
```

```{r plotting showing the figure}
ggplot(totalSteps, aes(x=tot)) +
        geom_histogram(fill= "red", binwidth=1000) +
        labs(title = "Histogram of Total Number of Daily Steps", x = "Steps", y = "Frequency")
```

### 3. Calculate and report the mean and median of the total number of steps taken per day

```{r calculating mean and median}
summary(totalSteps$tot)
```

===========================================================================

## What is the average daily activity pattern?

### 1. Make a time series plot of the 5-minute interval (x-axis) 

```{r code making plot}
# recode interval to factor variable 
activity$interval <- factor(activity$interval)

# calculate average steps per interval across days
avePattern <- ddply(activity, "interval", summarise, average=mean(steps, na.rm=T))
head(avePattern, 10)
str(avePattern)
```

``` {r code saving figure to folder, results = "hide", echo = FALSE}
jpeg("figure/figure2.jpeg", height=600, width=800, quality=8000)
plot(avePattern$interval, avePattern$average, 
     xlab="interval", ylab="average of steps", type="l",
     main="Average Daily Activity Pattern")
lines(avePattern$average)
dev.off()
```

```{r code showing the figure}
plot(avePattern$interval, avePattern$average, 
     xlab="interval", ylab="average of steps", type="l",
     main="Average Daily Activity Pattern")
lines(avePattern$average)
```

### 2. Which 5-minute interval contains the maximum number of steps?

```{r code arranging the data in decreasing order}
max <- arrange(avePattern, average, decreasing = TRUE)
head(max)
```

===========================================================================

## Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset 

```{r}
summary(activity$steps)
```

### 2. Impute missing values using the median

```{r imputing with median}
# Check the column names containing missing observations 
list_na <- colnames(activity)[apply(activity, 2, anyNA)]
list_na

# Replace missing values with the mean
activity_impute_median <- data.frame(
        sapply(activity,
                function(x) ifelse(is.na(x),
                                   median(x, na.rm = TRUE),
                                   x)))

```

### 3. Create a new dataset 

```{r}
summary(activity_impute_median$steps)
```

### 4. Histogram of the total number of steps using imputed data

```{r}
totalSteps2 <- ddply(activity_impute_median, "date", summarise, tot=sum(steps))
head(totalSteps2)
```

```{r plotting saving data, results="hide", echo=FALSE}
jpeg("figure/figure3.jpeg", height=600, width=800, quality=8000)
ggplot(totalSteps2, aes(x=tot)) +
        geom_histogram(fill= "blue", binwidth=1000) +
        labs(title = "Histogram of Total Number of Daily Steps, Imputed", x = "Steps", y = "Frequency")
dev.off()
```

```{r plotting showing figure}
ggplot(totalSteps2, aes(x=tot)) +
        geom_histogram(fill= "blue", binwidth=1000) +
        labs(title = "Histogram of Total Number of Daily Steps, Imputed", x = "Steps", y = "Frequency")
```

```{r calculating mean and median of the new data set}
summary(totalSteps2$tot)
```

#### with NA data, mean and median are 10766 and 10755

#### with imputed data (filled with median), the new mean and median are 9354 and 10395

===========================================================================

## Are there differences  between weekdays and weekends?

### 1. Create a new factor variable indicating whether a given date is a weekday or weekend day.

```{r creating weekday indicator}
# change date to date format
activity_impute_median$date <- as.Date(activity$date, format = "%Y-%m-%d")
activity_impute_median$day <-weekdays(activity_impute_median$date)

# binary indicator
activity_impute_median$day <-gsub("Monday|Tuesday|Wednesday|Thursday|Friday", "weekday", activity_impute_median$day) 
activity_impute_median$day <-gsub("Saturday|Sunday","weekend", activity_impute_median$day)  
table(activity_impute_median$day)
str(activity_impute_median)
```

### 2. Figure of daily pattern by weekday/weekend

```{r code plotting}
activity_impute_median$interval <- factor(activity_impute_median$interval)
activity_impute_median$day <- factor(activity_impute_median$day)

# summarize by weekday/weekend
act1 <- filter(activity_impute_median, activity_impute_median$day=="weekday")
aveP1 <- ddply(act1, "interval", summarise, average=mean(steps))
head(aveP1, 10)

act2 <- filter(activity_impute_median, activity_impute_median$day=="weekend")
aveP2 <- ddply(act2, "interval", summarise, average=mean(steps))
head(aveP1, 10)
```


```{r plotting by weekday/weekend}
par(mfrow=c(2, 1))
plot(aveP1$interval, aveP1$average, 
     xlab="interval", ylab="average of steps", type="l",
     main="Average Daily Activity Pattern - Weekdays")
lines(aveP1$average, col="blue")

plot(aveP2$interval, aveP2$average, 
     xlab="interval", ylab="average of steps", type="l",
     main="Average Daily Activity Pattern - Weekends")
lines(aveP2$average, col="red")
```

```{r plotting saving figure, results="hide", echo=FALSE}
jpeg("figure/figure4.jpeg", height=600, width=800, quality=8000)
par(mfrow=c(2, 1))
plot(aveP1$interval, aveP1$average, 
     xlab="interval", ylab="average of steps", type="l",
     main="Average Daily Activity Pattern - Weekdays")
lines(aveP1$average, col="blue")

plot(aveP2$interval, aveP2$average, 
     xlab="interval", ylab="average of steps", type="l",
     main="Average Daily Activity Pattern - Weekends")
lines(aveP2$average, col="red")
dev.off()
```

